---
title: "Trend-stationary vs difference-stationary time series"
author: "Nayef Ahmad"
date: "2022-07-18" 
output: 
   github_document: 
     toc: true
     number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview 
Someone gives you data that looks like this, and asks, "I know the OLS line has +ve slope, but is that a *real* trend, or is it just a fluke?" What do you do? 

[image]

First, give kudos for acknowledging a fundamental statistical idea: the difference between systematic vs "random" effects (purely due to sampling variability). They are right to hesitate before over-interpreting the OLS line. 

Next, remember that substantive questions have to be translated into statistical models, and there can be more than one way of doing so. Your client is strongly hinting that you should interpret the slope coefficient in this model: 

[image]

However, this is not the only model that can account for this data. At least one other model is worth considering: 

[image]

This model says that there is no systematic trend. Instead, there is a systematic *dependence structure* across values of the series. 


## References
1. Nelson & Kang, *Pitfalls in the use of time as an explanatory variable in regression*, 1984. 

2. Durlauf & Phillips, *Trends versus random walks in time series analysis*, 1988. 


# Libraries
```{r}
library(forecast)

```


# Example 
```{r}
set.seed(2021)

series_length <- 25
ar_coefficient <- .6

y <- arima.sim(series_length, 
               model = list(ar = ar_coefficient))
time <- 1:series_length
plot(time, y, type = "l", lwd = 2, main="Some time series data")

fit <- lm(y~time)
abline(fit, col = "blue")

```


Now let's see if we can recover those parameters: 

```{r}
auto.arima(y)

```



