---
title: "Inadequacy of t-tests for time series data"
author: "Nayef Ahmad"
date: "2022-03-04" 
output: 
   github_document: 
     toc: true
     number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview 

It's commonly known that violation of the independence assumption is a problem for the t-test. One common specific way in which this violation can occur is when there is dependence across time/sequence for a set of data. Here we show that this can lead to serious inflation of the Type 1 error rate. 

- todo: show that the Type 1 error inflation is worse when the AR coefficient is higher. 

Reference: 

- [Speegle & Clair, Section 8.5.4](https://mathstat.slu.edu/~speegle/_book/HTCI.html)


# Libraries 
```{r}
library(forecast)

```


# Single t-test

```{r}
layout.matrix <- matrix(c(1, 2, 1, 3), nrow = 2, ncol = 2)
layout(mat = layout.matrix)


time_series <- arima.sim(model = list(ar = .5), n = 100)
plot(time_series)
fit <- lm(as.vector(time_series) ~ c(1:length(time_series)))
abline(fit, col = "blue")

acf(time_series)
pacf(time_series)

sample_mean <- mean(time_series)

```

Here we have an `AR(1)` model with no "intercept" or starting term. By definition, it is weakly stationary, with mean equal to zero. That is, $\mu = 0$. 

The sample mean here is `r sample_mean`. 

We can do a t-test to assess the null hypothesis $H_0: \mu = 0$ vs alternative hypothesis $H_a: \mu  != 0$. In this case, we know that in reality $\mu = 0$. 

```{r}
t_test_p_value <- t.test(time_series, mu=0)$p.value
alpha <- .05
reject_null <- t_test_p_value < alpha
```


The t-test gives a p-value of `r t_test_p_value`, which in this case means `reject_null` = `r reject_null`. 


# Simulation of a large number of t-tests

To get a better sense of the performance of the t-test here, let's simulate a large number of null conditions, and see what proportion are wrongly rejected. 

```{r}
alpha <- 0.05

sim_data <- replicate(10000, {
  x <- arima.sim(model = list(ar = .5), n = 200)
  t.test(x, mu = 0)$p.value
  
})

prop_false_rejection <- mean(sim_data < alpha)
```

The Type 1 error rate is the proportion of time the p-value is less than `alpha` under the null hypothesis. 

**Here, the value of the Type 1 error rate is `r prop_false_rejection`. This is much higher than `alpha`, which is the "advertised" Type 1 error rate.** 


# Comparison using normally distributed data 

```{r}
sim_data <- replicate(10000, {
  x <- rnorm(100, mean = 0)
  t.test(x, mu = 0)$p.value
  
})

prop_false_rejection_normal <- mean(sim_data < .05)
```

Here the t-test performs well. The Type 1 error rate is `r prop_false_rejection_normal`


# Working with residuals from fitted ARIMA model 

We can use `auto.arima()` to fit an ARIMA model that accounts for the time series structure in our data. If the model accounts for most of the structure, then the residuals should be close to white noise - that is, independent and identically distributed. Thus, the t-test should work well on the residuals. 

```{r}
layout.matrix <- matrix(c(1, 2, 1, 3), nrow = 2, ncol = 2)
layout(mat = layout.matrix)


time_series <- arima.sim(model = list(ar = .5), n = 200)
model <- auto.arima(time_series, stepwise = FALSE, approximation = FALSE)

plot(time_series)
lines(model$fitted, col = "blue")

if (is.na(model$coef["intercept"])) {
  intercept <- 0
} else {
  intercept <- model$coef["intercept"]
}


hist(model$residuals)

shifted_residuals <- model$residuals + intercept
hist(shifted_residuals)


alpha <- 0.05
t_test_p_value <- t.test(shifted_residuals, mu = 0)$p.value
reject_null <- t_test_p_value < alpha
```

In this case, the p-value is `r t_test_p_value`, and `reject_null` is `r reject_null`


## Simulation over many iterations 

Let's repeat the experiment over a large number of iterations. 

```{r}
alpha <- 0.05

sim_data <- replicate(1000, {
  x <- arima.sim(model = list(ar = .5), n = 200)
  
  model <- auto.arima(x, stepwise = TRUE, approximation = TRUE)
  
  if (is.na(model$coef["intercept"])) {
    intercept <- 0
  } else {
    intercept <- model$coef["intercept"]
  }
  
  shifted_residuals <- model$residuals + intercept
  
  t.test(shifted_residuals, mu = 0)$p.value
  
})

prop_false_rejection <- mean(sim_data < alpha)

```

**The Type 1 error rate after removing some time series structure is `r prop_false_rejection`. Although this is still inflated, it is an improvement over the uncorrected t-test result.** 